A Simple Ratio for Estimating the Effective Reproduction Value of an Epidemic

### 2021-01-16 docs/manuscript/manuscript.Rmd
improved description of how simple ratio equation is derived

As described by previous publications [@wallinga_how_2007], we can define a function $n(a)$ that defines the expected number of new infections created today (time $t$) by a single person who was infected $a$ days ago (at time $t-a$). Therefore, to calculate $N_a$ (the total expected number of new infections created today by all individuals who became infected $a$ days ago), we multiply $n(a)$ by the number of new infections created $a$ days ago ($I_{t-a}$).

```{=tex}
\begin{equation}
  N_a = I_{t-a}n(a)
  (\#eq:eq2)
\end{equation}
```
Then to determine the total expected number of new infections today ($I_t$), we sum $N_a$ for all values of $a$ (i.e. sum up the expected number of new infections created by each infected individuals based on the age of their infection). If we treat $a$ and $t$ instead of discrete days, we integrate over all values of $a$ to produce Equation \@ref(eq:renewal-eq).

```{=tex}
\begin{equation} 
  I_t = \int_a^\infty{I_{t-a}n(a)}da
  (\#eq:renewal-eq)
\end{equation}
```
Next we can substitute the generation interval function ($g(a)$) into Equation \@ref(eq:renewal-eq) by recognizing that the generation interval distribution is equal to $n(a)$ normalized by the total number of infections a single infected person is expected cause (i.e. $R_t$).

```{=tex}
\begin{equation} 
  g(a) = \frac{n(a)}{R_t}
  (\#eq:generation-interval-eq)
\end{equation}
```
```{=tex}
\begin{equation} 
  I_t = \int_a^\infty{I_{t-a}\frac{g(a)}{R_t}}da
  (\#eq:sub-in-rt-1)
\end{equation}
```
```{=tex}
\begin{equation} 
  \frac{I_t}{R_t} = \int_a^\infty{I_{t-a}g(a)}da
  (\#eq:sub-in-rt-2)
\end{equation}
```
```{=tex}
\begin{equation} 
  R_t = \frac{I_t}{\int{I_{t-a}g(a)}da}
  (\#eq:sub-in-rt-3)
\end{equation}
```
If we make the assumption that there is minimal variance in the generation interval distribution (i.e. $g(a)$ equals approximately 1 when $a$ equals the mean generation interval ($d$) and $g(a)$ equals approximately zero for all other values of $a$), then Equation \@ref(eq:renewal-eq) can be simplified to:

@@

To assess accuracy of $R_t$ estimations, we used Susceptible-Exposed-Infectious-Resolved (SEIR) model to simulate epidemics and then compare how accurately the two methods can estimate the $R_t$ values in the epidemics. We simulated `r simulation_parameters$Nsims` epidemics in a population of `r addUnits(N)` individuals. We varied the lengths of latent and infectiousness periods with each simulation. In each simulation, the population underwent changes in the $R_t$ value representing changes in population behavior (e.g. "lockdown" period or lifting of a lockdown). \cref{fig:sim-results}**A** shows the true $R_t$ and the estimated $R_t$ values for each day in one such simulated epidemic. A full visualization of the all the simulations can be seen in (\cref{fig:all-simulations)).

Over the `r simulation_parameters$Nsims` simulation, the median estimation error of the daily $R_t$ by the simple ratio method was `r error_median` (90% range of `r error_quantile_025` to `r error_quantile_975`) (\cref{fig:sim-results)**B**). The simple ratio has a tendency to over-estimate the $R_t$, especially as the true value of $R_t$ increases above 1.5 (\cref{fig:sim-results)**C**). This is consistent with previous mathematical analysis of this simplified method of estimating $R_t$ [@wallinga_how_2007].

@@

Often however, the the mean and standard deviation of the generation interval is typically not known exactly. For instance, for the COVID-19 the mean generation interval estimates, with 95% credible intervals, include 3.95 (3.01 - 4.91) and 5.20 (3.78 - 6.78); and estimates of the standard deviation of the generation interval include 1.51 (0.74 - 2.97) and 1.72 (0.91 - 3.93) [@ganyani_estimating_2020]. Therefore, we compared the accuracy of the two methods if the standard deviation of the generation interval is mis-specified as three days longer. Under these conditions, the simple ratio actually performs slightly better than than the Cori method (\cref{fig:estimation-inaccParams)).

@@

We next compared $R_t$ estimates from the Simple Ratio and Cori et al for using data from the current COVID-19 pandemic. We estimated $R_t$ in four nations using a generation interval with mean of `r COVID19_params$mean_gi` and standard deviation of `r COVID19_params$sd_gi`[@ganyani_estimating_2020]. \cref{fig:real-world-data)**A** shows the results for the month of October 2020. Visual inspection shows that the two methods of estimating $R_t$ have very good concordance. Across all nations over the course of the pandemic, the median difference in estimation between the two methods was `r estimation_diff_median` (95% range of `r estimation_diff_quantile_025` to `r estimation_diff_quantile_975`) (\cref{fig:real-world-data)**B**).

>>>>

As described by previous publications [@wallinga_how_2007], we can define a function $n(a)$ that describes the expected number of new infections created today (time $t$) by a single person who was infected $a$ days ago (at time $t-a$). Therefore, to calculate $I_{ta}$ (the expected number of new infections created today by all individuals who became infected $a$ days ago), we multiply $n(a)$ by the number of new infections created $a$ days ago ($I_{t-a}$).

```{=tex}
\begin{equation}
  I_{ta} = I_{t-a}n(a)
  (\#eq:eq2)
\end{equation}
```

Then to determine the total expected number of new infections today ($I_t$), we sum $I_{ta}$ for all values of $a$. In other words, $I_t$ can be calculated by taking all individuals infected prior to $t$, calculating the expected number of new infections each one will produce based on the age of their infection, and summing all those values together. If we treat $a$ as continuous variables instead of discrete days, we can integrate over all values of $a$, as shown in \cref{eq:renewal-eq-1}.

```{=tex}
\begin{equation} 
  I_t = \int_a^\infty{I_{ta}}da
  (\#eq:renewal-eq-1)
\end{equation}
```
```{=tex}
\begin{equation} 
  I_t = \int_a^\infty{I_{t-a}n(a)}da
  (\#eq:renewal-eq-2)
\end{equation}
```

Next we can substitute the generation interval function ($g(a)$) into \cref{eq:renewal-eq-2} by recognizing that the generation interval distribution is equal to $n(a)$ normalized by the total number of infections a single infected person is expected cause (i.e. $R_t$).

```{=tex}
\begin{equation} 
  g(a) = \frac{n(a)}{R_t}
  (\#eq:generation-interval-eq)
\end{equation}
```
```{=tex}
\begin{equation} 
  I_t = \int_a^\infty{I_{t-a}g(a)R_t}da
  (\#eq:sub-in-rt-1)
\end{equation}
```
```{=tex}
\begin{equation} 
  \frac{I_t}{R_t} = \int_a^\infty{I_{t-a}g(a)}da
  (\#eq:sub-in-rt-2)
\end{equation}
```
```{=tex}
\begin{equation} 
  R_t = \frac{I_t}{\int{I_{t-a}g(a)}da}
  (\#eq:sub-in-rt-3)
\end{equation}
```
If we make the assumption that there is minimal variance in the generation interval distribution (i.e. $g(a)$ equals approximately 1 when $a$ equals the mean generation interval ($d$) and $g(a)$ equals approximately zero for all other values of $a$), then Equation \cref{eq:sub-in-rt-3} can be simplified to:

@@

To assess accuracy of $R_t$ estimations, we used Susceptible-Exposed-Infectious-Resolved (SEIR) model to simulate epidemics and then compare how accurately the two methods can estimate the $R_t$ values in the epidemics. We simulated `r simulation_parameters$Nsims` epidemics in a population of `r addUnits(N)` individuals. We varied the lengths of latent and infectiousness periods with each simulation. In each simulation, the population underwent changes in the $R_t$ value representing changes in population behavior (e.g. "lockdown" period or lifting of a lockdown). \cref{fig:sim-results}**A** shows the true $R_t$ and the estimated $R_t$ values for each day in one such simulated epidemic. A full visualization of the all the simulations can be seen in (\cref{fig:all-simulations}).

Over the `r simulation_parameters$Nsims` simulation, the median estimation error of the daily $R_t$ by the simple ratio method was `r error_median` (90% range of `r error_quantile_025` to `r error_quantile_975`) (\cref{fig:sim-results}**B**). The simple ratio has a tendency to over-estimate the $R_t$, especially as the true value of $R_t$ increases above 1.5 (\cref{fig:sim-results}**C**). This is consistent with previous mathematical analysis of this simplified method of estimating $R_t$ [@wallinga_how_2007].

@@

Often however, the the mean and standard deviation of the generation interval is typically not known exactly. For instance, for the COVID-19 the mean generation interval estimates, with 95% credible intervals, include 3.95 (3.01 - 4.91) and 5.20 (3.78 - 6.78); and estimates of the standard deviation of the generation interval include 1.51 (0.74 - 2.97) and 1.72 (0.91 - 3.93) [@ganyani_estimating_2020]. Therefore, we compared the accuracy of the two methods if the standard deviation of the generation interval is mis-specified as three days longer. Under these conditions, the simple ratio actually performs slightly better than than the Cori method (\cref{fig:estimation-inaccParams}).

@@

We next compared $R_t$ estimates from the Simple Ratio and Cori et al for using data from the current COVID-19 pandemic. We estimated $R_t$ in four nations using a generation interval with mean of `r COVID19_params$mean_gi` and standard deviation of `r COVID19_params$sd_gi`[@ganyani_estimating_2020]. \cref{fig:real-world-data}**A** shows the results for the month of October 2020. Visual inspection shows that the two methods of estimating $R_t$ have very good concordance. Across all nations over the course of the pandemic, the median difference in estimation between the two methods was `r estimation_diff_median` (95% range of `r estimation_diff_quantile_025` to `r estimation_diff_quantile_975`) (\cref{fig:real-world-data}**B**).
